---
title: "Kaggle - Titanic"
subtitle: "Step 1: Data Management"
author: "Michael Foley"
date: "4/27/2020"
output: 
  html_document:
    theme: flatly
    toc: true
    highlight: haddock
    fig_width: 9
    fig_caption: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The [Titanic: Machine Learning from Disaster](https://www.kaggle.com/c/titanic) competition challenges participants to predict the survivorship (1 = survived, 0 = perished) of passengers aboard the April 1912 maiden voyage of the Titanic.  In total, 1,502 of the 2,224 passengers and crew perished (68%). The training (*n* = 891) and test (*n* = 418) datasets comprise only a fraction of the total (59%). The data includes 10 features. Competitors build a predictive model with the training dataset, then apply the model to the test dataset to produce a submission file consisting of `PassengerId` and `Survived` (1|0).  Kaggle evaluates submissions on *accuracy*.

This file handles the data management, EDA, and feature engineering.  I got a lot of ideas from [here](https://www.kaggle.com/mrisdal/exploring-survival-on-the-titanic).


# Setup

```{r message=FALSE}
library(tidyverse)
library(caret)
library(mfstylr)
library(mice)
```


# Load Data

Here are the [column definitions](https://www.kaggle.com/c/titanic/data) in the dataset:

Variable | Definition
---|---
PassengerId	| 
Survived | Survival (0 = No, 1 = Yes)
Pclass | Passenger class (1 = 1st, 2 = 2nd, 3 = 3rd)
Name |				
Sex	| 		
Age	|	Passenger age (years)
SibSp	| # of siblings / spouses aboard the Titanic
Parch	| # of parents / children aboard the Titanic
Ticket | Ticket number
Fare | Passenger fare ($s)
Cabin	| Cabin number
Embarked | Port of Embarkation (C = Cherbourg, Q = Queenstown, S = Southampton)

```{r warning=FALSE, message=FALSE}
train <- read_csv("./train.csv") 
test <- read_csv("./test.csv")
full <- bind_rows(train, test)
train_index <- c(1:891)

glimpse(full)
```

Change the character vars to factors.

```{r}
full$Survived <- factor(full$Survived)
full$Pclass <- factor(full$Pclass)
full$Sex <- factor(full$Sex)
full$Embarked <- factor(full$Embarked)
```


# Feature Engineering

At this point, I know the full dataset is 1,309 x 12, including 10 feature columns.  How are we with data types and missingness?

```{r}
skimr::skim(full)
```

Cabin is 78% `NA`, `Age` is 20% `NA`, and there is 2 `NA` for `Embarked` and 1 `NA` for `Fare`.

## Impute NA: Embarked

Passengers `Embarked` from three ports (C = Cherbourg, Q = Queenstown, S = Southampton).  Two passengers have `NA`.

```{r}
full %>% filter(is.na(Embarked))
```

Both traveled without parents or siblings, so we can't try to pair them up with anyone.  We could guess by passenger class and fair.  Both were first class passengers, paying $80 for cabin B28.

```{r}
full %>% filter(Cabin == "B28")
```

```{r}
full %>%
  ggplot(aes(x = as.factor(Pclass), y = Fare, color = Embarked)) + geom_boxplot() +
  labs(title = "Embarked by Fare and Class")
```

Seems like port `C` is a good guess.  Could be port `S` too.  What does `mice()` say?

```{r}
set.seed(2020)
mice_obj <- mice(
  full[, !names(full) %in% c("PassengerId", "Name", "Ticket", "Cabin")],
  method = "rf"
)
mice_cmpl <- complete(mice_obj)
mice_cmpl[c(62, 830), ] %>% select(Embarked, everything())
```

Let's go with `S` then.

```{r}
full$Embarked <- mice_cmpl$Embarked
```


## Impute NA: Fare

One passenger has no listed fare.

```{r}
full %>% filter(is.na(Fare))
```

`PassengerId` 1044 was a third-class traveler from Southampton.  

```{r}
med <- full %>% filter(Embarked == "S" & Pclass == 3) %>% select(Fare) %>% 
  map(median, na.rm = TRUE) %>% unlist()
full %>%
  ggplot(aes(x = Embarked, y = Fare, color = Pclass)) + 
  geom_boxplot(na.rm = TRUE) +
  scale_y_continuous(labels = scales::dollar) +
  theme_mf() +
  labs(
    y = "",
    title = "Fare by Embarked and Class",
    subtitle = paste0(
      "Median fare for Southampton third class is ", scales::dollar(med), ".")
  )
```

The median fare for Southampton third-class was $8.05.  What did `mice()` say?

```{r}
mice_cmpl[1044, ] %>% select(Fare, everything())
```

$9.22 is reasonably close, so let's go with that.

```{r}
full$Fare <- mice_cmpl$Fare
```


## Impute NA: Age

20% of passengers have no listed `Age`.  I have no good ideas for imputation, so I'll just go with mice.

```{r}
data.frame(full = full$Age, mice = mice_cmpl$Age) %>%
  pivot_longer(cols = c("full", "mice"), names_to = "Data", values_to= "Age") %>%
  ggplot(aes(x = Age, color = Data)) + 
  geom_density(na.rm = TRUE) +
  theme_mf() +
  labs(title = "Age Distribution", color = "")
```

Mice looks great.

```{r}
full$Age <- mice_cmpl$Age
```


## Name -> Surname, Title, Name

`Name` includes a title and surname.  The title may be predictive of survivorship.  The surname may be useful for constructing family units (do families sink or swim together?).  I learned from other analyses, and supported by [Encycolpedia Titanica](https://www.encyclopedia-titanica.org/titanic-deckplans/), that the first letter of the cabin refers to the deck.  I'll parse these two fields first.

Separate `Name` into `Surname`, `Title`, `Name`.

```{r}
full <- full %>% 
   separate(Name, ", ", into = c("Surname", "Name")) %>%
   separate(Name, "\\. ", into = c("Title", "Name"), extra = "merge")

full %>% count(Title, sort = TRUE) 
```

There are 18 titles. If I turn them all into dummies, most will have near-zero variance.  From the documentation in `caret::nearZeroVar()`, a near zero var is one that is <10% unique and the frequency ratio to the second-most common value is > 95/5 = 19.  For dummy vars, the second-most common value is just the negation, so I can calculate both these metrics for Title's values.

```{r}
check_nzv <- function(dat, col) {
  dat %>% count(!!sym(col), sort = TRUE) %>%
    mutate(
      unique = round(n / sum(n), 2),
      ratio = n / (sum(n) - n),
      ratio = round(if_else(ratio < 1, 1 / ratio, ratio), 0),
      nzv = if_else(unique < 0.10 & ratio > 19, "Yes", "No")
    )
}
full %>% check_nzv("Title")
```

Only the first 3 have-non-zero variance.  "Master" nearly makes the cut.  The top 4 comprise 98% of the rows.  "Mlle" (mademoiselle) and "Ms" both mean "Miss".  "Mme" means "Mrs".  I'll merge these synonyms.  The rest can go into "Other".

```{r}
full <- full %>%
  mutate(
    Title = case_when(
      Title %in% c("Mlle", "Ms") ~ "Miss",
      Title == "Mme" ~ "Mrs",
      Title %in% c("Miss", "Mrs", "Mr", "Master") ~ Title,
      TRUE ~ "Other"
    ),
    Title = parse_factor(Title)
  )
full %>% check_nzv("Title")
```

So basically this variable classifies men, women, boys, and girls. I wonder how valuable that is given the `Sex` and `Age` variables already in the datas set.  `Surname` has 875 unique values, so the average party size is a little less than 2.  Not sure what I can do with `Surname`.

```{r}
n_distinct(full$Surname)
```

## Cabin -> Deck

`Cabin` contains both the deck [A-G] + [T] (Tank Top?) and cabin number (deck descriptions [here](https://jamescameronstitanic.fandom.com/wiki/Titanic_Decks)).  There are a few examples of multiple cabins, and in a few of these instances the multiple cabins are on multiple decks.  

```{r}
unique(full$Cabin)
```

I will asssume the first listed cabin contains the deck number.

```{r}
full$Deck <- map_chr(full$Cabin, str_sub, start = 1, end = 1) %>% factor()
full %>% count(Deck)
addmargins(table(full$Deck, full$Pclass, useNA = "ifany"))
```

There are 8 decks listed. One fellow seems to have been assigned to the tank top, meaning he's sleeping with the ship boilers.  

```{r}
full %>% filter(Deck == "T")
```

First class passenger too - that seems wrong.  The decks are labeled in descending level above water, so the highest deck is "A" and it would have been the most exclusive.  Let's verify that.

```{r}
full %>% count(Deck, Pclass) %>%
  ggplot(aes(x = Deck, y = n, fill = fct_rev(as.factor(Pclass)))) + geom_col() +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(legend.position = "top") +
  labs(title = "Passengers by Deck", fill = "Class")
```

Decks A-C are all first-class passengers, D has a few second-class, etc.  Makes sense (except for our first class passenger sleeping with the ship boilers).  Lots of nulls.  Seems like a good guess that the third-class passengers are on deck F or G.  How about average ticket fare by deck?

```{r}
full %>% group_by(Deck, Pclass) %>% 
  summarize(n = n(), mean_fare = mean(Fare, na.rm = TRUE)) %>%
  ggplot(aes(x = Deck, y = mean_fare, color = fct_rev(as.factor(Pclass)), size = n)) +
  geom_jitter() +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(legend.position = "top") +
  labs(title = "Mean Ticket Price by Deck", color = "Class")
```

Looks like deck A was pretty good value.  I better look into that more closely.

```{r}
full %>% filter(Deck == "A") %>% select(-Survived, -Surname, -Name, -Ticket, -Cabin)
```

Nothing unusual, except that Passenger ID 807's fair was $0.  Anyone else with $0 fare?

```{r}
full %>% filter(Fare == 0)
```

17 passengers with $0 fare.  All were male. Of the 15 in the training set, 14 perished.  Passenger class varied from 1 - 3.  All embarked from Southampton.  Could they have been crew?

Okay, my strategy is to un-assign the deck from Mr. Tank Top, let mice impute the NAs, then decide if I should collapse factors F and G. 

```{r}
full[340, ]$Deck <- NA
full$Deck = fct_drop(full$Deck)

set.seed(2020)
mice_obj <- mice(
  full[, !names(full) %in% c("PassengerId", "Name", "Ticket", "Cabin")],
  method = "rf"
)
mice_cmpl <- complete(mice_obj)
```

```{r}
addmargins(table(mice_cmpl$Deck, mice_cmpl$Pclass, useNA = "ifany"))
```

Deck G is sparse, so let's go ahead and collapse the level into F.

```{r}
full$Deck <- mice_cmpl$Deck
full$Deck <- fct_collapse(full$Deck, F = "G")
addmargins(table(full$Deck, full$Pclass, useNA = "ifany"))
```


## Family structure

The number of siblings plus spouse on board `SibSp`, and number of children and parents on board `Parch` present interesting possibilities.  Does party size improve survival odds?  Would a husband and wife fair better with or without a child present?  

### Family Size

```{r fig.height=6.5}
full[train_index, ] %>%
  mutate(fam_size = SibSp + Parch + 1) %>%
  count(Title, fam_size, Survived) %>%
  ungroup() %>% 
  group_by(Title, fam_size) %>%
  mutate(Pct = if_else(Survived == 1, n / sum(n), as.numeric(NA))) %>%
  ggplot(aes(x = fam_size, y = n, fill = as.factor(Survived), label = scales::percent(Pct))) + 
  geom_col() +
  geom_text(size = 3, na.rm = TRUE) +
  scale_fill_brewer(palette = "Paired") +
  scale_x_discrete(limits = c(1:11)) +
#  theme_minimal() +
  theme(legend.position = "top") +
  labs(title = "Survival by Title", fill = "Survived") +
  facet_wrap(~Title, nrow = 3)
```

I'll make family size a factor variable with bins 1, 2, and 3+

```{r}
full <- full %>%
  mutate(Fsize = factor(if_else(SibSp + Parch + 1 >=3, 3, SibSp + Parch + 1)))
```


### Role (Mom, Dad, Child)

```{r}
full %>%
  mutate(
    Role = case_when(
      Sex == "female" & Age > 50 & Parch >= 1 ~ "Grandmother",
      Sex == "female" & Age >= 20 & Age <= 50 & Parch >= 1 ~ "Mother",
      Sex == "female" & Age < 20 & Parch >= 1 ~ "Daughter",
      Sex == "male" & Age <= 50 & Parch >= 1 ~ "Father",
      Sex == "male" & Age > 50 & Parch >= 1 ~ "Grandfather",
      Sex == "male" & Age < 20 & Parch >= 1 ~ "Son",
      TRUE ~ "Alone"
    )
  )
```


```{r}
skimr::skim(full)
```


# Create Test and Train




# Save Work

```{r}
#save(d_raw_1, 
#     file = "./titanic_01.RData")
```

