---
title: "Kaggle - Titanic"
subtitle: "Step 1: Data Management"
author: "Michael Foley"
date: "4/27/2020"
output: 
  html_document:
    theme: flatly
    toc: true
    highlight: haddock
    fig_width: 9
    fig_caption: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The [Titanic: Machine Learning from Disaster](https://www.kaggle.com/c/titanic) competition challenges participants to predict the survivorship (1 = survived, 0 = perished) of passengers aboard the April 1912 maiden voyage of the Titanic.  In total, 1,502 of the 2,224 passengers and crew perished (68%). The training (*n* = 891) and test (*n* = 418) data sets comprise only a fraction (59%) of the total. The data includes 10 features. Competitors build a predictive model with the training data set, then apply the model to the test data set to produce a submission file consisting of `PassengerId` and `Survived` (1|0).  Kaggle evaluates submissions on *accuracy*.

This file handles the data management and feature engineering.  I got a lot of ideas from [here](https://www.kaggle.com/mrisdal/exploring-survival-on-the-titanic).


# Setup

```{r message=FALSE}
library(tidyverse)
library(caret)
library(mfstylr)
library(mice)
library(janitor)
```


# Load Data

Here are the train/test data set column definitions from [Kaggle](https://www.kaggle.com/c/titanic/data).

Variable | Definition
---|---
PassengerId	| 
Survived | Survival (0 = No, 1 = Yes)
Pclass | Passenger class (1 = 1st, 2 = 2nd, 3 = 3rd)
Name |				
Sex	| 		
Age	|	Passenger age (years)
SibSp	| # of siblings / spouses aboard the Titanic
Parch	| # of parents / children aboard the Titanic
Ticket | Ticket number
Fare | Passenger fare ($s)
Cabin	| Cabin number
Embarked | Port of Embarkation (C = Cherbourg, Q = Queenstown, S = Southampton)

```{r warning=FALSE, message=FALSE}
train <- read_csv("./train.csv") 
test <- read_csv("./test.csv")
full <- bind_rows(train, test)
train_index <- c(1:891)

glimpse(full)
```


# Feature Engineering

At this point, I know the full data set is 1,309 x 12, including 10 feature columns.  How are the data types and missingness?

```{r}
skimr::skim(full)
```

`Survived`, `Pclass`, `Sex`, and `Embarked` should be factors.

```{r}
full$Survived <- factor(full$Survived)
full$Pclass <- factor(full$Pclass)
full$Sex <- factor(full$Sex)
full$Embarked <- factor(full$Embarked)
```

`Cabin` is 77% null, `Age` is 20% null, and there are 2 nulls for `Embarked` and 1 for `Fare`.  I'll take care of NAs next.


## Fare -> Fare per Ticket

`Fare` is the price of the ticket, but there can be multiple passengers per ticket.  The per-person price is `Fare` divided number of passengers on the ticket.  Let's untangle that by 1) creating a variable `TicketN` for the number of passengers on the ticket, and 2) creating a variable `FarePerPass` for the per-passenger price.

### Passengers per Ticket

Create new predictor, `TicketN`, for number of passengers per ticket.

```{r}
full <- full %>% group_by(Ticket) %>% mutate(TicketN = n()) %>% ungroup()

full %>% ggplot(aes(x = TicketN)) + geom_histogram(bins = 30) +
  scale_x_continuous(n.breaks = 15) +
  labs(title = "Passenger Count Distribution") +
  theme_minimal()
```

### Impute NA: Fare

Define `FarePerPass` as `Fare` by `TicketN`.  First, however, I need to deal with the passenger ticket with missing `Fare`.  

```{r}
full %>% filter(is.na(Fare))
```

`PassengerId` 1044 was a third-class traveler boarding at Southampton.  What is the typical fare per passenger for in this profile?

```{r}
full %>% 
  filter(Embarked == "S" & Pclass == 3) %>% 
  mutate(FarePerPass = Fare / TicketN) %>% 
  select(FarePerPass) %>% 
  summary()
```

The median fare per passenger was $7.796. 

```{r}
full[full$PassengerId == 1044, ]$Fare <- 7.796
```


### Fare per Passenger

Calculate `FarePerPass` now.

```{r}
full$FarePerPass <- full$Fare / full$TicketN
```

Let's see what the distribution of fares look like now.

```{r}
p1 <- full %>% ggplot(aes(x = FarePerPass, color = Pclass)) + geom_density() +
  theme_minimal() + 
  theme(legend.position = "top") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "FarePerPass Distribution")
p2 <- full %>% ggplot(aes(x = Pclass, y = FarePerPass, color = Pclass)) + 
  geom_boxplot() + geom_jitter(alpha = 0.6) +
  theme_minimal() + 
  theme(legend.position = "top") + 
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "FarePerPass Box Plot")
gridExtra::grid.arrange(p1, p2, nrow = 1)
```

First class tickets can cost up to $125 per passenger!  Even more surprising is the tickets costing $0 per passenger.

## Fare

17 passengers paid $0 for their ticket.  All were male. Of the 15 in the training set, 14 perished. 

```{r}
full %>% filter(Fare == 0) %>% arrange(Name)
```

After a little research [here](https://www.encyclopedia-titanica.org/titanic-passenger-list/), it appears that [nine of the passengers](https://titanic.fandom.com/wiki/Guarantee_Group) who paid $0 were from the Guarantee Group from Harland & Wolff, the shipbuilding firm.  They were regarded as crew.  Of the others, [Reuchlin](https://www.encyclopedia-titanica.org/titanic-victim/johan-george-reuchlin.html) (PassengerId 823) was sort of an employee. So were [Fry, Harrison, and Ismay](https://www.encyclopedia-titanica.org/titanic-survivor/j-bruce-ismay.html) (816, 264, and 1264).  Perhaps the best way to deal with these passengers is to create a new employee variable, `Employee` (1 = yes, 0 = no), set their `FarePerPass` to `NA`, and impute a "market-rate" fare.

```{r}
full <- full %>% 
  mutate(
    Employee = if_else(Fare == 0, 1, 0),
    FarePerPass = if_else(FarePerPass == 0, as.numeric(NA), FarePerPass)
  )
```

I'll use Mice to impute.  You might think the response variable `Survived` would be useful to include, but 1) it feels cicular to use it, and 2) it is often missing.  Anyway, I experimented with and without it, and leaving it out yielded more accurate imputations.

```{r}
set.seed(2020)
mice_obj <- mice(
  full[, !names(full) %in% c("PassengerId", "Name", "Ticket", "Cabin", "Survived")],
  method = "rf"
)
mice_cmpl <- complete(mice_obj)
```

```{r}
data.frame(full = full$FarePerPass, mice = mice_cmpl$FarePerPass) %>%
  replace_na(list(full = 0)) %>%
  pivot_longer(cols = c("full", "mice"), names_to = "Data", values_to= "Fare") %>%
  ggplot(aes(x = Fare, color = Data)) + 
  geom_density(na.rm = TRUE) +
  theme_minimal() + 
  scale_x_continuous(labels = scales::dollar) +
  theme(legend.position = "top") +
  labs(title = "Fare Distribution", color = "")
```

Those 17 fit in invisibly.  Assign the imputed values.  Recalculate `Fare` too.

```{r}
full$FarePerPass <- mice_cmpl$FarePerPass
full$Fare <- full$FarePerPass * full$TicketN
```

One more look at `FarePerPass`.

```{r}
p1 <- full %>% ggplot(aes(x = FarePerPass, color = Pclass)) + geom_density() +
  theme_minimal() + 
  theme(legend.position = "top") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "FarePerPass Distribution")
p2 <- full %>% ggplot(aes(x = Pclass, y = FarePerPass, color = Pclass)) + 
  geom_boxplot() + geom_jitter(alpha = 0.6) +
  theme_minimal() + 
  theme(legend.position = "top") + 
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "FarePerPass Box Plot")
gridExtra::grid.arrange(p1, p2, nrow = 1)
```


## Impute NA: Embarked

Passengers `Embarked` from three ports (C = Cherbourg, Q = Queenstown, S = Southampton).  Two passengers have `NA`.

```{r}
full %>% filter(is.na(Embarked))
```

They traveled on the same ticket, so they probably embarked from the same port.  They paid $40 per passenger.  Where did other passengers fitting this profile embark?

```{r}
plot_prof <- function(hline, subtitle){
  full %>%
    filter(!is.na(Embarked)) %>%
    ggplot(aes(x = Embarked, y = FarePerPass, color = Pclass)) + 
    geom_boxplot(na.rm = TRUE) +
    geom_hline(yintercept = hline, linetype = 2) +
    scale_y_continuous(labels = scales::dollar) +
    theme_mf() +
    scale_color_brewer(palette = "Paired") +
    labs(
      y = "Fare",
      title = "Passenger Profiles",
      subtitle = subtitle
    )
}
plot_prof(40, "First Class passengers paying $80/ticket probably embarked from Cherbourg")
```

Seems like port Cherbourg is a good guess, but it could be any of them.  What did `mice()` predict?

```{r}
mice_cmpl[c(62, 830), ] %>% select(Embarked, everything())
```

`mice()` thinks Chebourg too.

```{r}
full$Embarked <- mice_cmpl$Embarked
```


## Impute NA: Age

20% of passengers have no listed `Age`.  I have no good ideas for imputation, so I'll just go with mice.

```{r}
data.frame(full = full$Age, mice = mice_cmpl$Age) %>%
  pivot_longer(cols = c("full", "mice"), names_to = "Data", values_to= "Age") %>%
  ggplot(aes(x = Age, color = Data)) + 
  geom_density(na.rm = TRUE) +
  theme_mf() +
  labs(title = "Age Distribution", color = "")
```

Mice looks great.

```{r}
full$Age <- mice_cmpl$Age
```


## Name -> Surname, Title, Name

`Name` includes a title and surname.  The title may be predictive of survivorship.  The surname may be useful for constructing family units (do families sink or swim together?), but I'm not sure how to go about doing that, so I'll focus on title.  I learned from other analyses, and supported by [Encycolpedia Titanica](https://www.encyclopedia-titanica.org/titanic-deckplans/), that the first letter of the cabin refers to the deck.  I'll parse these two fields first.

Separate `Name` into `Surname`, `Title`, `Name`.

```{r}
full <- full %>% 
   separate(Name, ", ", into = c("Surname", "Name")) %>%
   separate(Name, "\\. ", into = c("Title", "Name"), extra = "merge")

full %>% count(Title, sort = TRUE) 
```

There are 18 titles. If I turn them all into dummies, most will have near-zero variance.  From the documentation in `caret::nearZeroVar()`, a near zero var is one that is <10% unique and the frequency ratio to the second-most common value is > 95/5 = 19.  For dummy vars, the second-most common value is just the negation, so I can easily calculate both these metrics.

```{r}
check_nzv <- function(dat, col) {
  dat %>% count(!!sym(col), sort = TRUE) %>%
    mutate(
      unique = round(n / sum(n), 2),
      ratio = n / (sum(n) - n),
      ratio = round(if_else(ratio < 1, 1 / ratio, ratio), 0),
      nzv = if_else(unique < 0.10 & ratio > 19, "Yes", "No")
    )
}
full %>% check_nzv("Title")
```

Only the first 3 have-non-zero variance.  "Master" nearly makes the cut.  The top 4 comprise 98% of the rows.  "Mlle" (mademoiselle) and "Ms" both mean "Miss".  "Mme" means "Mrs".  I'll merge these synonyms.  The remaining items will comprise a very small proportion of values, so even grouping them into "Other" will make the factor NZV.  I'll take some guesses for them.

```{r}
full <- full %>%
  mutate(
    Title = case_when(
      Title %in% c("Mlle", "Ms") ~ "Miss",
      Title == "Mme" ~ "Mrs",
      Title %in% c("Miss", "Mrs", "Mr", "Master") ~ Title,
      Title %in% c("Dr", "Rev", "Major", "Col", "Capt") & Sex == "male" ~ "Mr",
      Title %in% c("Dr") & Sex == "female" ~ "Mrs",
      Title == "Lady" & Age >= 30 ~ "Mrs",
      Title %in% c("Don", "Sir", "Jonkheer") ~ "Mr",
      Title %in% c("the Countess", "Dona") ~ "Mrs",
      TRUE ~ Title
    ),
    Title = parse_factor(Title)
  ) 
full %>% check_nzv("Title")
```

So basically this variable classifies men, women, boys, and girls. (I wonder how valuable that is given the `Sex` and `Age` variables already in the data set).  `Surname` has 875 unique values, so the average party size is a little less than 2.  Not sure what I can do with `Surname`.

```{r}
n_distinct(full$Surname)
```

## Cabin -> Deck

`Cabin` contains both the deck [A-G, T] (T = Tank Top?) and cabin number (deck descriptions [here](https://jamescameronstitanic.fandom.com/wiki/Titanic_Decks)).  There are a few rows with multiple cabins, and in a few of these instances the multiple cabins are on multiple decks.  

```{r}
unique(full$Cabin)
```

I will asssume the *first* listed cabin contains the deck number.

```{r}
full$Deck <- map_chr(full$Cabin, str_sub, start = 1, end = 1) %>% factor()
```

Here is the distribution of values, grouped by passenger class.

```{r}
full %>%
  tabyl(Deck, Pclass) %>% 
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting() %>%
  adorn_ns() %>%
  adorn_title("combined") %>%
  knitr::kable()
```

There are 8 decks listed. One fellow seems to have been assigned to the tank top, meaning he's sleeping with the ship boilers.  

```{r}
full %>% filter(Deck == "T")
```

First class passenger too - that seems wrong.  What else... The decks are labeled in descending level above water, so the highest deck is "A" and it would have been the most exclusive.  That is evident from the table above - decks A, B, and C are 100% first class.  Lots of missing values though.  Seems like a good guess that the third-class passengers are on deck F or G.  How about average ticket fare by deck?

```{r}
full %>% 
  ggplot(aes(x = fct_explicit_na(Deck), y = FarePerPass, 
             color = as.factor(Pclass))) +
  geom_boxplot() +
  scale_color_brewer(palette = "Paired") +
  theme_minimal() +
  scale_y_continuous(labels = scales::dollar) +
  theme(legend.position = "top") +
  labs(title = "FarePerPass by Deck", color = "Class", x = "Deck", y = "Fare per Passenger")
```

Okay, so what am I going to do with `Deck`?  First, I'll un-assign the deck from Mr. Tank Top since that just seems wrong, and I can't have a factor variable level with one row.  Then I'll let mice impute the NAs.  After that, I'll decide if I should collapse factors F and G since G was pretty sparse. 

```{r}
full[340, ]$Deck <- NA
full$Deck = fct_drop(full$Deck)

set.seed(2020)
mice_obj <- mice(
  full[, !names(full) %in% c("PassengerId", "Name", "Ticket", "Cabin", "Survived")],
  method = "rf"
)
mice_cmpl <- complete(mice_obj)
```

```{r}
addmargins(table(mice_cmpl$Deck, mice_cmpl$Pclass, useNA = "ifany"))
```

```{r}
mice_cmpl %>%
  tabyl(Deck, Pclass) %>% 
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting() %>%
  adorn_ns() %>%
  adorn_title("combined") %>%
  knitr::kable()
```

Deck G is still sparse, so let's go ahead and collapse the level into F.

```{r}
full$Deck <- mice_cmpl$Deck
full$Deck <- fct_collapse(full$Deck, F = "G")
```

Oh, where did Mr. Tank Top go?

```{r}
full %>% filter(PassengerId == 340) %>% select(Deck, everything())
```

Deck B!  


## Family structure

The number of siblings plus spouse on board, `SibSp`, and number of parents and children on board, `Parch` present interesting possibilities.  Does *total* family size improve survival odds? 

```{r}
plot_surv_grp <- function(dat, x_grp, facet_grp){
  dat %>%
  group_by(!!sym(facet_grp), !!sym(x_grp)) %>%
  summarize(n = n(), pct_surv = sum(if_else(Survived == "1", 1, 0)) / n()) %>%
  ggplot(aes(x = !!sym(x_grp), y = pct_surv, fill = n)) + 
  geom_col(position = position_dodge()) +
  scale_fill_steps(low = "goldenrod1", high = "goldenrod4") +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ Title)
}
full[train_index, ] %>%
  mutate(fam_size = as.factor(SibSp + Parch + 1)) %>% plot_surv_grp("fam_size", "Title")
```

In general, survival odds fall with increasing family size, although the n-size is pretty small for families over three.  Men buck the trend a little - their surivorship is relatively low when traveling alone.  I'll make family size a factor variable, `Fsize`, with bins 1, 2, and 3+.  For modeling, it probably makes sense to have an interaction variable for `Title*Fsize`.

```{r}
full <- full %>%
  mutate(
    Fsize = if_else(SibSp + Parch + 1 >=3, 3, SibSp + Parch + 1),
    Fsize = factor(Fsize, levels = c(1, 2, 3), labels = c("1", "2", "gte3"))
  )

full[train_index, ] %>%
  plot_surv_grp("Fsize", "Title")
```


## Net Survivors

Is the survivorship of the other passengers on the ticket predictive of your survivorship?  Create feature `TicketSurvN`.

```{r}
full <- full %>%
  group_by(Ticket) %>%
  mutate(TicketNetSurv = sum(Survived == "1", na.rm = TRUE) - sum(Survived == "0", na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    TicketNetSurvOth = TicketNetSurv + case_when(Survived == "1" ~ -1,
                                                 Survived == "0" ~ +1,
                                                 is.na(Survived) ~ 0)
  )
```


# Remove Unnecesary Vars

The text fields are the chaffe, so discard them.

```{r}
full <- full %>% 
  select(-Surname, -Name, -Ticket, -Cabin, -Fare) %>%
  select(PassengerId, Survived, sort(names(.))) 
```

One final look.

```{r}
skimr::skim(full)
```

# Save Work

```{r}
save(full, train_index, file = "./titanic_01.RData")
```

